---
title: "Calculate Sun Angle"
author: "Sebastian Di Geronimo"
date: 2025-06-05
format: html
editor: source
---

# 1.0 ---- Summary of Document ----

This file calculates the nautical dawn and dusk based on location and day.

The time and location of sample is then compared to nautical dawn and dusk to 
determine if the sample was taken during day time, night time, or cerpuslar 
dya/night (i.e. 1 hour buffer around nautical dawn/dusk).

# 2.0 ---- Setup ----


## 2.1 Load Libraries

```{r setup, include=FALSE}
if (!nzchar(system.file(package = "librarian"))) 
  install.packages("librarian")

librarian::shelf(
  quiet = TRUE,
  librarian, conflicted, ggplot2, tibble, tidyr, readr, purrr, dplyr, stringr,
  forcats, lubridate, glue, fs, magrittr, here,
  
  # additional
  suncalc
)

conflicts_prefer(
  dplyr::filter(), 
  dplyr::select()
  )
  

```

## 2.2 Load Dataset

Load microscopy metadata

```{r load-data}
zoo_time <-
  here("data") %>%
  dir_ls(regexp = "obis_merg") %>%
  read_csv(show_col_types = FALSE) %>%
  select(cruise_id, site, station, mesh, date_time, date, time_gmt, lat_in,
         lon_in) %>%
  distinct() %>%
  mutate(
    date = as_date(date),
    sun_angle = pmap(
      list(.x = date, .y = lat_in, .z = lon_in),
      \(.x, .y, .z) getSunlightTimes(
        date = .x, lat = .y, lon = .z,
        keep = c("nauticalDawn", "nauticalDusk")
      )
    )
  ) %>%
  unnest(sun_angle, names_sep = "_") %>%
  mutate(
     day_night = case_when(
      between(
        date_time,
        sun_angle_nauticalDawn + hours(1),
        sun_angle_nauticalDusk - hours(1)
      ) ~ "day",
      between(
        date_time,
        sun_angle_nauticalDawn,
        sun_angle_nauticalDawn + hours(1)
      ) ~ "crepuscular dawn",
      between(
        date_time,
        sun_angle_nauticalDusk - hours(1),
        sun_angle_nauticalDusk
      ) ~ "crepuscular dusk",
      .default = "night"
    )
  ) %T>% 
  print()

zoo_time %>%
  count(day)

zoo_time %>%
  count(mesh, day)
zoo_time %>%
  count(station, day)
```


```{r metadata-full}
if (dir_exists(here("~", "Box", "IMaRS Zooplankton data"))) {
  meta_path <- here(
    "~", "Box", "IMaRS Zooplankton data", "cruise_logsheets",
    "meta_combined.csv"
  )
} else {
  meta_path <- rstudioapi::selectFile()
}

meta_full <-
  meta_path %>%
  read_csv(show_col_types = FALSE) %>%
  select(cruise_id:lon_in) %T>%
  print()


meta_sun <-
  meta_full %>%
  mutate(
    sun_angle = pmap(
      list(.x = date, .y = lat_in, .z = lon_in),
      \(.x, .y, .z) getSunlightTimes(
        date = .x, lat = .y, lon = .z,
        keep = c("nauticalDawn", "nauticalDusk")
      )
    )
  ) %>%
  unnest(sun_angle, names_sep = "_") %>%
  mutate(
    day_night = case_when(
      between(
        date_time,
        sun_angle_nauticalDawn + hours(1),
        sun_angle_nauticalDusk - hours(1)
      ) ~ "day",
      between(
        date_time,
        sun_angle_nauticalDawn,
        sun_angle_nauticalDawn + hours(1)
      ) ~ "crepuscular dawn",
      between(
        date_time,
        sun_angle_nauticalDusk - hours(1),
        sun_angle_nauticalDusk
      ) ~ "crepuscular dusk",
      .default = "night"
    )
  ) %T>%
  print()

```

```{r}
meta_sun %>%
  count(day_night)

meta_sun %>%
  count(mesh_size_um, day_night)

meta_sun %>%
  count(station, day_night)
```
# 3.0 Save Day/Night Info

```{r save-sun}
# ---- save data_set
meta_sun %>%
  write_csv(
    x    = .,
    file = here(rstudioapi::selectDirectory(), "meta_day_night.csv"),
    na = ""
  )

```

